Model-order reduction of non-linear time dependent neutronics via POD-Galerkin projection and DEIM

### 2022-03-16 paper/interactnlmsample.tex
Jeremy's comments

\usepackage{algorithm,algorithmic}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{caption}
\usepackage{threeparttable}
\usepackage{tikz}
\usepackage{multirow}
\usepackage{booktabs,tabularx}
\usepackage{subcaption}
\DeclareMathOperator{\argmaxH}{argmax} 
\DeclareMathOperator*{\argmin}{argmin}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\newcommand\abs[1]{\left|#1\right|}

@@

\maketitle

\begin{abstract}
Presented here is an application of model-order reduction to non-linear neutronic transients, where reduced-order models (ROMs) are generally sought to provide a rapid and accurate approximation of a particular simulation solution.
The ROM was built with the use of the POD-Galerkin projection method, where the main contribution of this work is the employment of the matrix version of the Discrete Empirical Interpolation Method (MDEIM) to improve the ROM efficiency.
The MDEIM is used to approximate the projection of the problem operator that otherwise is required to be computed at each non-linear iteration within each time step, increasing the ROM cost.
By using the MDEIM, an offline-online ROM decomposition was obtained, which is favored from a computational perspective since the expensive projection operation can take place only once in the offline stage.
The online stage, which involves solving a linear system  at each time step, depends only on the reduced dimension and is therefore cheap, making it possible to increase the solution time resolution, and accordingly accuracy, without introducing significant cost.
By applying the method to the LRA benchmark, speeds up that range from 6 to 40 was achieved depending on the full-order model (FOM) spatial fidelity.
The maximum relative error in the assemblies power was in the order of $1\times10^{-5}$\%.
Moreover, the ROM was parameterized by using a POD-greedy sampling and a case study was tested where the kinetic data. i.e, the precursors decay constant and delayed neutron fraction, are assumed to be the parameters of interest.
For the parameterized ROM, the maximum relative error in the assembly power was  \% ., which indicates that the developed ROM can be reliably used in parametric and uncertainty quantification studies.
The computational gain from the MDEIM-ROM was compared with that of a direct projection ROM and showed that employing the MDEIM is computationally advantageous and becomes even more appreciable with higher-dimensional problems.
The method was implemented in the open-source deterministic transport code Detran where only access to the system operator was needed to construct the ROM, and hence the method is considered trivially invasive and could be implemented in modern deterministic transport codes.

@@

\section{Introduction}

Modeling and simulation of reactor physics is crucial to understand nuclear reactors behavior under different conditions and scenarios and hence ensure reliable design and well established safety limits.
The mathematical models governing the reactor physics are described by a set of equations that is challenging to solve analytically due to the geometric complexity of nuclear reactors except for simplified cases.
Alternatively, these systems are solved by employing numerical discretization that rely on discretizing the system along each independent variable and the solution is provided only on these discrete points.
To improve the method accuracy, the system needs to be solved over a very fine mesh.
Although this results in a more reliable solution, it leads to a growing space dimension and accordingly large computational cost.
This problem of the computational cost becomes inevitably prohibitive with applications that require repetitive execution of the simulation such as uncertainty quantification, and design optimization.
In such studies, the high-fidelity model has to be solved at multiple parameter points which could be very large.
Hence, surrogate models that are cheap, yet accurate, are sought to provide a rapid and reliable approximation of the solution of these systems.
The premise is that this lower-order model will capture the essential features of the system  and preserve the input-output mapping as much as possible.

In the application of our interest here, neutronic transients, the mathematical models are based on the diffusion approximation and the Boltzmann transport equation.
For realistic modeling, these neutron transport models need to be coupled with different physics models to account for their feedback.
For example, thermal-hydraulics feedback plays an essential role in reactor operation.
Changes in the fuel temperature alter the cross sections due to Doppler broadening \cite{stacey2018nuclear}, which in turn impacts the neutron population and hence the fission rate and the rate of heat generation.
Multi-physics modeling accounts for these different feedback mechanisms; however, it poses a challenge in terms of the computational cost,
since it entails solving coupled systems for different fields such as neutron flux, temperature, and flow rate.
Moreover, the neutron transport (or diffusion) equation is no longer a linear one due to the indirect dependence of the cross section on the neutron flux.
Two approaches are used to solve such systems.
The most commonly used one is a serial coupling of two or more separate codes in such a way that the output of one at a given time step, such as the flux distribution, is used as an input to the other, e.g., the temperature (which can be used then to update the cross section and so on).
This procedure is commonly referred to as operator splitting or explicit coupling.
One drawback of operator splitting is that time integration requires small steps to avoid stability issues, thus leading to high computational time. 
Alternatively, tight coupling in time between the physics can be maintained by using an implicit scheme that solves a non-linear iteration in each time step, also increasing the computational time.
A modern algorithm such as Jacobian free Newton-Krylov (JFNK) \cite{gaston2009parallel} can be used to perform this implicit coupling.
Another popular approach is to use fixed-point iteration, with an implicit time scheme to iterate between difference physics within a time step until convergence is satisfied, which is computationally taxing.
The aim of this work is to use a reduce-order model to overcome the high computational expense of these non-linear models.
In particular, the ROM was developed using the POD-Galerkin projection.
The method belongs to the projection-based family, and can be described as a physics-driven method since it requires knowledge of the original governing equation as will be shown in Sec.\ref{sec:POD}.
In a previous work \cite{elzohery2021modeling}, we applied the method to time dependent diffusion equation where there was no coupling with other models and hence the application was linear.
In this study, the developed ROM was extended to treat the non-linearity induced by multi-physics modeling.
To improve the computational efficiency of the developed ROM, a hyper-reduction technique known as the Discrete Empirical Interpolation Method (DEIM) was used in conjunction of the POD-Galerkin projection.
The primary purpose of using the DEIM is to obtain an offline-online decomposition of the ROM to avoid repetitive projection of the problem operator at each iteration within the time step.
The offline stage occurs only once and it comprises the expensive step of the ROM, where the online stage is cheaper and involves solving a reduced number of equation at each time step.
The LRA benchmark was used as a demonstrative application where the diffusion equation is coupled with a simplified thermal feedback model.

@@

Consider the following dynamic system that results from discretizing a partial differential equation using a numerical scheme such as the finite difference method:
\begin{equation}
	\frac{d\mathbf{y}(t)}{dt} = \mathbf{A}\mathbf{y}(t) + \mathbf{f}(\mathbf{y}(t)),  \quad\quad \mathbf{y(0) = \mathbf{y_0}} \, ,
	\label{eq:FOM}
\end{equation}
where $\mathbf{y} \in \mathbb{R}^{n}$ is the state variable and $\mathbf{A} \in \mathbb{R}^{n\times n}$ is the operator resulting from the system discretization.
In general, this method involves two main steps.
First, a subspace is obtained using the proper orthogonal decomposition, i.e, POD \cite{}, a widely used reduction technique.
Commonly, the POD subspace is generated using the snapshots method \cite{} by constructing a data matrix $\mathbf{Y}$ containing the system solutions at different time instances, then the singular value decomposition (SVD) is computed, i.e $\mathbf{Y} = \mathbf{U}\mathbf{\Sigma}\mathbf{V}$.
The POD subspace is the one spanned by the first $r$ vectors of $\mathbf{U}$, where $r \ll n$.
There is no general \textit{a priori} rule based on which the rank $r$ is chosen; however one can set a threshold $\epsilon$  such that
\begin{equation}
	\frac{\sum_{i=0}^r S_i}{\sum_{i=0}^m S_i} \ge \epsilon \, ,
\end{equation}
where S$i$ is the $i$th singular value.
Upon obtaining the reduced space, the state variable $\mathbf{y}$ is expanded in terms of its basis
\begin{equation}
	\mathbf{y} \approx \mathbf{U} \hat{\mathbf{y}}(t) \, ,
	\label{eq:galerkin expansion}
\end{equation}
where $\mathbf{\hat{y}(t)} \in \mathbb{R}^{r}$ is a vector of the temporal coefficients.
Second, the system is projected onto a POD subspace by multiplying the system on the left by $\mathbf{U}^T$ and making use of the orthogonality of the POD basis (i.e, $\mathbf{U}^T\mathbf{U} = \mathbf{I}$), we end up with 
\begin{equation}
	\frac{d\mathbf{\hat{y}}}{dt} = \mathbf{A}_r\mathbf{\hat{y}} + \mathbf{U}^T\mathbf{f}(\mathbf{U}\mathbf{\hat{y}}) \qquad \mathbf{\hat{y}}(0) = \mathbf{U}^T\mathbf{y}_0 \, ,
	\label{eq:ROM}
\end{equation}
where $\mathbf{A}_r \in \mathbb{R}^{r\times r} = \mathbf{U}^T\mathbf{A}\mathbf{U}.$
Thus, we are left with $r$ ODE's instead of $n$ ODE's, which makes the resulting ROM cheaper to solve, and once it is solved for $\mathbf{\hat{y}}$, the full order solution can be computed by Eq.~\ref{eq:galerkin expansion}.

@@

The method is a discrete variant of the Empirical Interpolation Method (EIM) \cite{barrault2004empirical} that was originally proposed to approximate
non-affine coefficient functions using a reduced-basis expansion to facilitate the online/offline decomposition in finite element methods.

@@

The reconstruction and the projection steps have cost dependent on the dimension of the system, i.e $n$.
The goal of using the DEIM is to avoid performing these operations at each time step and split the cost of the ROM into two steps,
offline step where the expensive operations dependent on $n$ take place, and online step that depends only on the reduced dimension.
The method relies on two main ingredients: (1) a reduced space that approximates the nonlinear function, and (2) a set of spatial interpolation indices.

@@

and the final function approximation is given by

@@

A key point relevant to the computational efficiency of the method is that the matrix $\mathbf{P}$ does not need to be constructed explicitly, since the term $\mathbf{P}^T\mathbf{f}$ extracts only the elements of $\mathbf{f}$ that correspond to the interpolation indices.

@@

Now, the projection can be performed by computing the term $\mathbf{U}^T \mathbf{\Psi}$ once in an offline stage, while in the online stage, the vector $\boldsymbol{\theta}$ is computed by solving the linear system of dimension $k$ in Eq.~\ref{eq:DEIM coefficients}.
Thus, the cost of the online computation depends on the reduced dimension $k$, not the original, high dimension $n$.

Although the problem tackled here does not include a non-linear term, the DEIM idea is exploited to approximate time-dependent operators to avoid computing the projected operator $\mathbf{A}_r(t)$ at each time step, which requires number of FLOPS dependent on $n$.
This is known as the matrix version of the DEIM, i.e, MDEIM and it was suggested in \cite{carlberg2012efficient, benner2015survey}.
To illustrate, consider the operator $\mathbf{A}(\boldsymbol{t}) \in \mathbb{R}^{n\times n}$.
The DEIM procedures can be used to approximate the operator by a summation of few time independent matrices, each weighted by a temporal coefficient.
First, the matrix can be cast in a vector form by stacking all of its nonzero elements in one vector $\mathbf{a}(t) \in \mathbb{R}^{n^2}$.
This vector can be treated in the same exact way as the function $\mathbf{f}(t)$ in the previous section, i.e, $\mathbf{a}(\mu) = \mathbf{\Psi}\boldsymbol{\theta}(t)$.
Snapshots of the serialized matrix are collected at different time instances and the reduced subspace $\mathbf{\Psi}$ and the interpolation indices are computed following algorithm \ref{alg:DEIM}.
Again, the corresponding coefficients are obtained by solving Eq.~\ref{eq:DEIM coefficients}. 
Next, each vector in $\mathbf{\Psi}$ can be mapped back to an $n \times n$ matrix, i.e, $\textbf{A}_i$.
Hence, the matrix is approximated as $\mathbf{\hat{A}}(t)=\sum_{i=0}^k \theta_i(t) \mathbf{A}_i$ and the projection term is computed as
\begin{equation}
	\mathbf{U}^T\mathbf{A}\mathbf{U} = \sum_{i=0}^k \mathbf{\theta}_i(t)\mathbf{U}^T\mathbf{A}_i\mathbf{U} \, .
	\label{Eq:deim matrix projection}
\end{equation}
As pointed out in Ref.\cite{negri2015efficient}, the Weyl\textendash Mirsky theorem \cite{weyl1912asymptotische} ensures that the singular values of the approximated matrix approach that of the original as $k$ increases 
\begin{equation}
	|\sigma_i - \hat{\sigma_i}  |\ \le \|\mathbf{A} - \hat{\mathbf{A}}\|_\text{F}
\end{equation}
where $\sigma_i$ and $\hat{\sigma_i}$ denote the $i$th singular value of $\mathbf{A}$ and $\hat{\mathbf{A}}$ respectively.
%Moreover, some properties of the original matrix such as positive definiteness can be preserved \cite{carlberg2015preserving}, if necessary, by imposing some constrains to the problem in Eq.~\ref{eq:DEIM coefficients}.

In the practical implementations, the method efficiency can be improved, as suggested in Ref. \cite{negri2015efficient}.
First, the matrix $\mathbf{A}$ is, generally, largely sparse, which leads to a snapshots matrix that is sparse, too.
Consequently, it is more convenient to store only the nonzero elements ($nz$) in $\mathbf{a}$, 
which should be appreciably more efficient in terms of memory and computational expense.
In addition, to obtain the expansion coefficients ($\boldsymbol{\theta}$), the term $\mathbf{P}^T\mathbf{a(t)}$ needs to be computed at each new parameter value.
In fact, this term only extracts the elements of $\mathbf{a(t)}$ that corresponds to the selected  interpolation indices.
In the offline stage, a mapping can be done between these interpolation indices and the corresponding row-column pairs in $\mathbf{A}$.
When the matrix is formed in the online stage, only these entries are evaluated by restricting a \textit{for loop} over these row-column pairs, leading to more reduction in the computational cost.

\noindent\begin{minipage}{\textwidth}
	\renewcommand\footnoterule{}
	\begin{algorithm}[H]
		\caption{DEIM}
		\label{alg:DEIM}
		\begin{algorithmic}[1]
			\REQUIRE Snapshots matrix: $\mathbf{X} = [\mathbf{f}_1, \mathbf{f}_2,  ...., \mathbf{f}_m] \in \mathbb{R}^{n\times m}$
			\ENSURE A reduced space $\mathbf{\Psi}$, set of indices $\mathcal{I}$
			\STATE $[\mathbf{\Psi}_1, \mathbf{\Psi}_2, ..., \mathbf{\Psi}_k ]$ = POD(${\mathbf{X}}, \epsilon$)\footnote{POD refers to Algorithm \ref{alg:pod}}
			\STATE $\mathcal{I}_1 = \argmaxH(\abs{\mathbf{\Psi}_1})$, $\mathcal{I}$ = $\{\mathcal{I}_1\}$
			\STATE $\mathbf{\Psi} = [\mathbf{\Psi}_1]$, $\mathbf{P} =[e_{I_1}]$
			\FOR {i=2 \TO i=k }
			\STATE Solve $(\mathbf{P}^T\mathbf{\Psi})\mathbf{c} = \mathbf{P}^T\mathbf{\Psi}_i$;
			\STATE $\mathcal{I}_i = \argmaxH{\abs{\mathbf{\Psi}_i - \mathbf{\Psi}\mathbf{c}}}$
			\STATE $\mathcal{I} = \mathcal{I} \cup \mathcal{I}_i$, $\mathbf{P} = \mathbf{P} \cup e_{I_i}$, ${\Psi = \Psi \cup \Psi_i}$
			\ENDFOR
		\end{algorithmic}
	\end{algorithm}
\end{minipage}

@@

A greedy sampling in the parameter space \cite{nguyen2009reduced} can be combined with POD to construct a global subspace that can be used in the framework of parametric ROMs.
The primary goal is to ensure that the parameter domain of interest (i.e, $\mathbb P$) is efficiently sampled and includes enough information such that it can be used or wide range of parameters including those of higher dimensions.
To construct this subspace, a training parameters dataset, $P_{train} \subset \mathbb P$, is prepared. 
The greedy subspace is initialized with the first $r_1$ POD basis computed from a randomly selected sample of the training data.
Next, this subspace is used to compute the ROM prediction error of the state variable for the rest of the training samples.
The sample at which the error is worst is selected, and the subspace is enriched by adding the first $r_1$ POD basis computed from the selected sample snapshots.
Selecting the samples with the worst errors ensures efficient sampling of the parameter space.
Intuitively, samples that are close to the previously chosen parameter points will have a smaller ROM error.
On the other hand, one expects that the ROM will fail to give an accurate prediction at samples that are far from the included points in the current subspace.
Thus, we keep adding new basis vectors until a maximum size of the subspace is reached or a user-specified error tolerance at a reference parameter point $\boldsymbol{\mu}_{\text{ref}}$ is met.
Note that each time a new basis vector is added, the subspace is orthonormalized.
Regarding the error criteria based on which a new sample is selected, there are two approaches. 
Ideally, the true error can be used leading to ``strong POD-greedy'' \cite{benner2017model}.
Alternatively, an error indicator can be used to alleviate the computational burden associated with evaluating the full-order-model. 
Here, we used the strong-POD greedy approach. 
The algorithm steps are shown in Algorithm \ref{alg:greedy pod}. 
The term $\Delta(\textbf{U},\boldsymbol{\mu}_\text{ref})$  is the error measure using the current reduced subspace at $\boldsymbol{\mu}_\text{ref}$, and $r$ is the maximum size of the greedy basis.

@@

		\STATE set n=0

		\WHILE {$\epsilon:= \Delta(\textbf{U}, \boldsymbol{\mu}_\text{ref}) > \epsilon_{tol}$ or $n < r$}

>>>>

\usepackage{algorithm,algorithmic}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{caption}
\usepackage{threeparttable}
\usepackage{tikz}
\usepackage{multirow}
\usepackage{booktabs,tabularx}
%\usepackage{subcaption}
\DeclareMathOperator{\argmaxH}{argmax} 
\DeclareMathOperator*{\argmin}{argmin}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\newcommand\abs[1]{\left|#1\right|}

@@

\maketitle
\begin{abstract}
Presented here is an application of model-order reduction to non-linear neutronic transients.
% where reduced-order models (ROMs) are generally sought to provide a rapid, yet accurate, approximation of a particular simulation response.
The effort described extends previous work\cite{elzohery2021modeling} aimed to develop and assess a ROM framework that can be used for different purposes, such as design optimization and propagating input parameter uncertainties like nuclear data.
Here, the target applications here is non-linear.
The ROM was built with the use of the POD-Galerkin projection method and the matrix version of the Discrete Empirical Interpolation Method (MDEIM) for improved efficiency.
The MDEIM is used to approximate the projection of the problem operator that otherwise must be computed at each non-linear iteration within each time step, which increases the ROM cost.
By using the MDEIM, an offline-online ROM decomposition was obtained, which is favored from a computational perspective since the expensive projection operation can take place only once in the offline stage.
The online stage, which involves solving a linear system  at each time step, depends only on the reduced dimension and is therefore cheap, making it possible to increase the solution time resolution, and accordingly accuracy, without introducing significant cost.
By applying the method to the LRA benchmark, speed ups from 6 to 40 were achieved depending on the full-order model (FOM) spatial fidelity.
The maximum relative error in the assemblies power was in the order of $1\times10^{-5}$\%.
Moreover, the ROM was parameterized by POD-greedy sampling, and a case study was performed where the kinetic data, i.e, the precursor decay constants and delayed-neutron fraction were assumed to be the model parameters of interest.
For the parameterized ROM, the maximum relative error in the assembly power was $1\times 10^{-3}$\%, which implies that the developed ROM can be reliably used to accelerate parametric and uncertainty quantification studies involving non-linear problems.
The computational gain from the MDEIM-ROM was compared with that of a direct projection ROM and showed that employing the MDEIM is computationally advantageous, and the advantage grows with dimension of the problem.
The method was implemented in the open-source deterministic transport code Detran\cite{roberts2014advanced} where only access to the system operator was needed to construct the ROM, and; hence, the method is trivially invasive and could easily be implemented in many modern deterministic diffusion codes.

@@

\section{Introduction}
Modeling and simulation of reactor physics is crucial to understanding nuclear reactors' behavior under different conditions and scenarios and hence ensure reliable designs and well-established safety limits.
The mathematical models governing reactor physics are described by a set of equations that are challenging to solve analytically due to the geometric complexity of nuclear reactors except for simplified cases.
Alternatively, these systems are solved by employing numerical discretization that relies on discretizing the system along each independent variable and the solution is provided only on these discrete points.
To improve the method accuracy, the system needs to be solved over a very fine mesh.
Although this results in a more reliable solution, it leads to a growing space dimension and accordingly large computational cost.
This problem of the computational cost becomes inevitably prohibitive with applications that require repetitive execution of the simulation such as uncertainty quantification, and design optimization.
In such studies, the high-fidelity model has to be solved at multiple parameter points which could be very large.
Hence, surrogate models that are cheap, yet accurate, are sought to provide a rapid and reliable approximation of the solution of these systems.
The premise is that this lower-order model will capture the essential features of the system and preserve the input-output mapping as much as possible.

In the application of our interest here, neutronic transients, the mathematical models are based on the diffusion approximation and the Boltzmann transport equation.
For realistic modeling, these neutron transport models need to be coupled with different physics models to account for their feedback.
For example, thermal-hydraulics feedback plays an essential role in reactor operation.
Changes in the fuel temperature alter the cross sections due to Doppler broadening \cite{stacey2018nuclear}, which in turn impacts the neutron population and hence the fission rate and the rate of heat generation.
Multi-physics modeling accounts for these different feedback mechanisms; however, it poses a challenge in terms of the computational cost,
since it entails solving coupled systems for different fields such as neutron flux, temperature, and flow rate.
Moreover, the neutron transport (or diffusion) equation is no longer a linear one due to the indirect dependence of the cross section on the neutron flux.
Two approaches are used to solve such systems.
The most commonly used one is a serial coupling of two or more separate codes in such a way that the output of one at a given time step, such as the flux distribution, is used as an input to the other, e.g., the temperature (which can be used then to update the cross section and so on).
This procedure is commonly referred to as operator splitting or explicit coupling.
One drawback of operator splitting is that time integration requires small steps to avoid stability issues, thus leading to high computational time. 
Alternatively, tight coupling in time between the physics can be maintained by using an implicit scheme that solves a non-linear iteration in each time step, also increasing the computational time.
A modern algorithm such as Jacobian free Newton-Krylov (JFNK) \cite{gaston2009parallel} can be used to perform this implicit coupling.
Another popular approach is to use fixed-point iteration, with an implicit time scheme to iterate between difference physics within a time step until convergence is satisfied, which is computationally taxing.
The aim of this work is to use a reduced-order model to overcome the high computational expense of these non-linear models.
In particular, the ROM was developed using the POD-Galerkin projection.
The method belongs to the projection-based family and can be described as a physics-driven method since it requires knowledge of the original governing equation as will be shown in Sec.\ref{sec:POD}.
In a previous work \cite{elzohery2021modeling}, we applied the method to the time-dependent diffusion equation where there was no coupling with other models and hence the application was linear.
In this study, the developed ROM was extended to treat the non-linearity induced by multi-physics modeling.
To improve the computational efficiency of the developed ROM, a hyper-reduction technique known as the Discrete Empirical Interpolation Method (DEIM) was used in conjunction with the POD-Galerkin projection.
The primary purpose of using the DEIM is to obtain an offline-online decomposition of the ROM to avoid repetitive projection of the problem operator at each iteration within the time step.
The offline stage occurs only once and comprises the expensive step of the ROM.
On the other hand, the online stage involves solving a reduced number of equations at each time step with a computational cost independent of the FOM dimension and hence cheaper.
The LRA benchmark was used as a demonstrative application where the diffusion equation is coupled with a simplified thermal feedback model.

@@

Consider the following dynamic system that results from the discretization a partial differential equation using, e.g, the finite-difference method:
\begin{equation}
	\frac{d\mathbf{y}(t)}{dt} = \mathbf{A}\mathbf{y}(t) + \mathbf{f}(\mathbf{y}(t)),  \quad\quad \mathbf{y(0) = \mathbf{y_0}} \, ,
	\label{eq:FOM}
\end{equation}
where $\mathbf{y} \in \mathbb{R}^{n}$ is the state variable and $\mathbf{A} \in \mathbb{R}^{n\times n}$ is the operator resulting from the system discretization.
In general, this method involves two main steps.
First, a subspace is obtained using proper orthogonal decomposition (POD)\cite{volkwein2011model}.
Commonly, the POD subspace is generated using the snapshots method \cite{sirovich1987turbulence} by constructing a data matrix $\mathbf{Y}$ containing the system solutions at different times of which the singular value decomposition (SVD) is computed, i.e, $\mathbf{Y} = \mathbf{U}\mathbf{\Sigma}\mathbf{V}^T$.
The POD subspace is the one spanned by the first $r$ vectors of $\mathbf{U}$, where $r \ll n$.
There is no general \textit{a priori} rule based on which the rank $r$ is chosen; however, one can set a threshold $\epsilon$  such that
\begin{equation}
	\frac{\sum_{i=0}^r S_i}{\sum_{i=0}^m S_i} \ge \epsilon \, ,
\end{equation}
where S$_i$ is the $i$th singular value and the singular values are the elements of the diagonal matrix $\mathbf{\Sigma}$.
Upon obtaining the reduced space, the state variable $\mathbf{y}$ is expanded in terms of its basis
\begin{equation}
	\mathbf{y} \approx \mathbf{U} \hat{\mathbf{y}}(t) \, ,
	\label{eq:galerkin expansion}
\end{equation}
where $\mathbf{\hat{y}(t)} \in \mathbb{R}^{r}$ is a vector of the temporal coefficients.
Second, the system is projected onto a POD subspace by multiplying the system on the left by $\mathbf{U}^T$ and making use of the orthogonality of the POD basis (i.e, $\mathbf{U}^T\mathbf{U} = \mathbf{I}$), which leads to
\begin{equation}
	\frac{d\mathbf{\hat{y}}}{dt} = \mathbf{A}_r\mathbf{\hat{y}} + \mathbf{U}^T\mathbf{f}(\mathbf{U}\mathbf{\hat{y}})\, ,  \qquad \mathbf{\hat{y}}(0) = \mathbf{U}^T\mathbf{y}_0 \, ,
	\label{eq:ROM}
\end{equation}
where $\mathbf{A}_r \in \mathbb{R}^{r\times r} = \mathbf{U}^T\mathbf{A}\mathbf{U}.$
Thus, we are left with $r$ ODE's instead of $n$ ODE's, which makes the resulting ROM cheaper to solve, and once it is solved for $\mathbf{\hat{y}}$, the full order solution can be computed by Eq.~\ref{eq:galerkin expansion}.

@@

The method is a discrete variant of the Empirical Interpolation Method (EIM) \cite{barrault2004empirical} that was originally proposed to approximate non-affine coefficient functions using a reduced-basis expansion to facilitate the online/offline decomposition in finite-element methods.

@@

The reconstruction and the projection steps have costs dependent on the dimension of the system $n$.
The goal of using the DEIM is to avoid performing these operations at each time step by splitting the ROM into two steps: an offline step in which the expensive operations dependent on $n$ are performed, and an online step in which operations depend only on the reduced dimension.
The method relies on two main ingredients: (1) a reduced space in which nonlinear function can be approximated, and (2) a set of spatial interpolation indices.

@@

and the final approximation is given by

@@

A key point relevant to the computational efficiency of the method is that the matrix $\mathbf{P}$ does not need to be constructed explicitly since the term $\mathbf{P}^T\mathbf{f}$ extracts only the elements of $\mathbf{f}$ that correspond to the interpolation indices.

@@

Now, the projection can be performed by computing the term $\mathbf{U}^T \mathbf{\Psi}$ once in an offline stage, while in the online stage, the vector $\boldsymbol{\theta}$ is computed by solving the linear system of dimension $k$ in Eq.~\ref{eq:DEIM coefficients}.
Thus, the cost of the online computation depends on the reduced dimension $k$, not the original, larger dimension $n$.

Although the problem tackled here does not include a non-linear term, the DEIM idea is exploited to approximate time-dependent operators to avoid computing the projected operator $\mathbf{A}_r(t)$ at each time step, which requires number of FLOPS dependent on $n$.
This is known as the matrix version of the DEIM, i.e, MDEIM and it was suggested in \cite{carlberg2012efficient, benner2015survey}.
To illustrate, consider the operator $\mathbf{A}(\boldsymbol{t}) \in \mathbb{R}^{n\times n}$.
The DEIM procedures can be used to approximate the operator by a summation of few time independent matrices, each weighted by a temporal coefficient.
First, the matrix can be cast in a vector form by stacking its elements in one vector $\mathbf{a}(t) \in \mathbb{R}^{n^2}$.
This vector can be treated in the same exact way as the function $\mathbf{f}(t)$ in the previous section, i.e, $\mathbf{a}(\mu) = \mathbf{\Psi}\boldsymbol{\theta}(t)$.
Snapshots of the serialized matrix are collected at different time instances and the reduced subspace $\mathbf{\Psi}$ and the interpolation indices are computed following algorithm \ref{alg:DEIM}.
Again, the corresponding coefficients are obtained by solving Eq.~\ref{eq:DEIM coefficients}. 
Next, each vector in $\mathbf{\Psi}$ can be mapped back to an $n \times n$ matrix, i.e, $\textbf{A}_i$.
Hence, the matrix is approximated as $\mathbf{\hat{A}}(t)=\sum_{i=0}^k \theta_i(t) \mathbf{A}_i$ and the projection term is computed as
\begin{equation}
	\mathbf{U}^T\mathbf{A}\mathbf{U} = \sum_{i=0}^k \mathbf{\theta}_i(t)\mathbf{U}^T\mathbf{A}_i\mathbf{U} \, .
	\label{Eq:deim matrix projection}
\end{equation}
As pointed out in Ref.\cite{negri2015efficient}, the Weyl\textendash Mirsky theorem \cite{weyl1912asymptotische} ensures that the singular values of the approximated matrix approach that of the original as $k$ increases 
\begin{equation}
	|\sigma_i - \hat{\sigma_i}  |\ \le \|\mathbf{A} - \hat{\mathbf{A}}\|_\text{F}
\end{equation}
where $\sigma_i$ and $\hat{\sigma_i}$ denote the $i$th singular value of $\mathbf{A}$ and $\hat{\mathbf{A}}$ respectively.
%Moreover, some properties of the original matrix such as positive definiteness can be preserved \cite{carlberg2015preserving}, if necessary, by imposing some constrains to the problem in Eq.~\ref{eq:DEIM coefficients}.

In practical implementations, the method efficiency can be improved, as suggested in Ref. \cite{negri2015efficient}.
First, the matrix $\mathbf{A}$ is, generally, largely sparse, which leads to a snapshots matrix that is sparse, too.
Consequently, it is more convenient to store only the nonzero elements ($nz$) in $\mathbf{a}$, 
which should be appreciably more efficient in terms of memory and computational expense.
In addition, to obtain the expansion coefficients ($\boldsymbol{\theta}$), the term $\mathbf{P}^T\mathbf{a(t)}$ needs to be computed at each new time step.
In fact, this term only extracts the elements of $\mathbf{a(t)}$ that corresponds to the selected  interpolation indices.
In the offline stage, a mapping can be done between these interpolation indices and the corresponding row-column pairs in $\mathbf{A}$.
When the matrix is formed in the online stage, only these entries are evaluated by restricting a \textit{for loop} over these row-column pairs, leading to more reduction in the computational cost.

For the rest of this paper, we will refer to the ROM with MDEIM as the MDEIM-ROM, while we refer to the model in Sec.~\ref{sec:POD} as the direct projection ROM.

\noindent\begin{minipage}{\textwidth}
	\renewcommand\footnoterule{}
	\begin{algorithm}[H]
		\caption{DEIM}
		\label{alg:DEIM}
		\begin{algorithmic}[1]
			\REQUIRE Snapshots matrix: $\mathbf{X} = [\mathbf{f}_1, \mathbf{f}_2,  ...., \mathbf{f}_m] \in \mathbb{R}^{n\times m}$
			\ENSURE A reduced space $\mathbf{\Psi}$, set of indices $\mathcal{I}$
			\STATE $[\mathbf{\Psi}_1, \mathbf{\Psi}_2, ..., \mathbf{\Psi}_k ]$ = POD(${\mathbf{X}}, \epsilon$)\footnote{POD refers to the POD procedures in Sec.~\ref{sec:POD}}
			\STATE $\mathcal{I}_1 = \argmaxH(\abs{\mathbf{\Psi}_1})$, $\mathcal{I}$ = $\{\mathcal{I}_1\}$
			\STATE $\mathbf{\Psi} = [\mathbf{\Psi}_1]$, $\mathbf{P} =[e_{I_1}]$
			\FOR {i=2 \TO i=k }
			\STATE Solve $(\mathbf{P}^T\mathbf{\Psi})\mathbf{c} = \mathbf{P}^T\mathbf{\Psi}_i$;
			\STATE $\mathcal{I}_i = \argmaxH{\abs{\mathbf{\Psi}_i - \mathbf{\Psi}\mathbf{c}}}$
			\STATE $\mathcal{I} = \mathcal{I} \cup \mathcal{I}_i$, $\mathbf{P} = \mathbf{P} \cup e_{I_i}$, ${\Psi = \Psi \cup \Psi_i}$
			\ENDFOR
		\end{algorithmic}
	\end{algorithm}
\end{minipage}

@@

A greedy sampling in the parameter space \cite{nguyen2009reduced} can be combined with POD to construct a global subspace that can be used in the framework of parametric ROMs.
The primary goal is to ensure that the parameter domain of interest (i.e, $\mathbb P$) is efficiently sampled and includes enough information such that it can be used or wide range of parameters including those of higher dimensions.
To construct this subspace, a training parameters dataset, $P_{train} \subset \mathbb P$, is prepared. 
The greedy subspace is initialized with the first $r_1$ POD basis computed from a randomly selected sample of the training data.
Using this subspace, the ROM prediction error of the state variable for the rest of the training samples is computed.
The sample at which the error is worst is selected, and the subspace is enriched by adding the first $r_1$ POD basis computed from the selected sample snapshots.
Selecting the samples with the worst errors ensures efficient sampling of the parameter space.
Intuitively, samples that are close to the previously chosen parameter points will have a smaller ROM error.
On the other hand, one expects that the ROM will fail to give an accurate prediction at samples that are far from the included points in the current subspace.
Thus, we keep adding new basis vectors until a maximum size of the subspace is reached or a user-specified error tolerance at a reference parameter point $\boldsymbol{\mu}_{\text{ref}}$ is met.
Note that each time a new basis vector is added, the subspace is orthonormalized.

Regarding the error criteria based on which a new sample is selected, there are two approaches. 
Ideally, the true error can be used leading to ``strong POD-greedy'' \cite{benner2017model}.
Alternatively, an error indicator can be used to alleviate the computational burden associated with evaluating the full-order-model. 
Here, we used the strong-POD greedy approach. 
The algorithm steps are shown in Algorithm \ref{alg:greedy pod}. 
The term $\Delta(\textbf{U},\boldsymbol{\mu}_\text{ref})$  is the error measure using the current reduced subspace at $\boldsymbol{\mu}_\text{ref}$, and $r$ is the maximum size of the greedy basis.

@@

		\STATE Set n=0; randomly choose $\boldsymbol{\mu}\in P_{train}$
		\STATE Evaluate the FOM at $\boldsymbol{\mu}$, $\textbf Y(\boldsymbol{\mu})$:= $\{\textbf{y}^1, \textbf{y}^2, ..... \textbf{y}^k\}$
		\STATE $\mathbf{U} := \{\zeta_1, \zeta_2, ......,\zeta_{r1}\} := \text{POD}(\textbf{Y}(\boldsymbol{\mu}))$
		\WHILE {$\epsilon:= \Delta(\textbf{U}, \boldsymbol{\mu}_\text{ref}) > \epsilon_{tol}$ or $n < r$}
